### 2025-05-01 docs/Homeownership_BankOfMomAndDad.tex
Added new retention model results

I first present empirical evidence indicating that parental wealth helps not only in the transition to ownership, but also in maintaining ownership. \textcolor{red}{bring in new empirical results here. Highlight that the results on behind and unemployment are new}. I show that, controlling for a wide set of household and parental characteristics, households with wealthier parents are less likely to fall behind on mortgage payments after becoming homeowners, even though they buy more expensive homes. Finally, using a variation of the event study in \cite{Chetty2007}, I show that while most households downsize their housing during unemployment spells, those with wealthy parents retain their current homes.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the homeownership rate among young households.

I then use the model to evaluate policies aiming to increase homeownership. I find that such policies disproportionately benefit households with wealthy parents. Relaxing mortgage constraints---for example, by lowering minimum down payments---strengthens the link between parental wealth and children's housing outcomes, since more households with wealthy parents are on the margin to become homeowners. Nonmortgage policies that lower barriers to entry, such as reducing prices or purchase costs, increase homeownership for all households, though disproportionately among those with wealthier parents. However, increasing liquidity by reducing sales costs weakens the link between parental wealth and children's housing outcomes, as the partial insurance provided by parental wealth becomes less important.

I then study how illiquidity affects the strategic behavior arising in the absence of commitment. I find that 27\% of young households prefer illiquid housing---due to sales costs---to liquid housing. These households have wealthy parents and use the sales costs as a commitment device to maintain low liquid wealth in the next period, effectively becoming ``wealthy hand-to-mouth.'' Lower liquid wealth implies higher marginal propensities to consume (MPC) and, hence, more parental transfers. As an illustrative example, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption (with a correspondingly high marginal value of wealth) or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home. This driver of preference for illiquidity differs from those studied in the literature, such as temptation or time inconsistency \citep{attanasio2024temptation,laibson1997golden}. Furthermore, parental wealth generating preference for illiquidity provides a novel explanation for the existence of high-MPC households \cite[see e.g.,][]{kaplan2022marginal}. That children with wealthy parents find illiquidity less problematic---or even prefer it---is consistent with the empirical results in \cite{choukhmane2023benefits}. They find that households with wealthy parents are more likely to invest in illiquid retirement accounts and less likely to make costly early withdrawals, even after controlling for a wide set of characteristics.

Finally, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults. I adapt the methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}, who study the racial wealth gap, and recalibrate the income process to capture racial differences. I find that parental transfers account for 7 percentage points (14\%) of the Black-White homeownership gap. For context, the Black-White homeownership gap has remained roughly constant since the passing of the Fair Housing Act in 1968.

@@

\subsubsection{Households with Wealthier Parents Retain Ownership}
While most research focuses on the transition into ownership, I now focus on the challenge of retaining it.

@@

\begin{table}[tb]
	\center 
	\begin{threeparttable}
	\caption{Model Estimation}\label{tab:esttable}
		\input{../tabfig/est/esttable.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} Standard errors calculated by estimating the model to 100 bootstrapped empirical samples.
		\end{flushleft}\end{footnotesize}}
\end{table}
\subsubsection{Nontargeted Moments}\label{sec:externalval}
To validate the model, I show that it also matches nontargeted moments.

@@

\begin{figure}
	\caption{Event Study: Housing Consumption at Unemployment by Parental Wealth}\label{fig:housinggrowthrates_model}
	\includegraphics[width=0.5\textwidth]{../tabfig/model_housinggrowthpoor_both}%
	\includegraphics[width=0.5\textwidth]{../tabfig/model_housinggrowthrich_both}

@@

Taking stock, this section shows that the model can match targeted moments regarding homeownership, transfers, and the transition to homeownership with realistic preference parameter values. The model also matches a set of untargeted moments. Perhaps more importantly, the model replicates the household-level associations between parental wealth and downsizing during income losses.

@@

The observed Black-White homeownership gap is $\frac{0.54 - 0.27}{0.54} = 50\%$, compared to 52\% in the model with altruism. Removing altruism reduces the gap to 45\%, implying that parental transfers account for 7 percentage points---or 14\%---of the gap. While Black and White households receive transfers at similar rates in the model, transfers are less influential for Black households, who are further from the ownership threshold and receive smaller amounts. These results highlight the importance of racial differences in parental income, which shape both wealth accumulation and transfer capacity.

@@

These results have several implications for the MPC literature. First, model-implied MPCs may be overstated when informal intra-family transfers are omitted, particularly for households with wealthy parents. Second, a common strategy to match high MPCs is to introduce households with low target wealth—e.g., through impatience \citep{aguiar2024hand}—but such behavior is observationally similar to that of households expecting altruistic support. Third, models that generate high MPCs by introducing illiquid assets with large return premia \citep{kaplan2022marginal} may overlook the fact that homeowners with wealthy parents have access to informal insurance, which lowers their effective MPC.

>>>>

I first present empirical evidence showing that parental wealth helps not only with the transition into ownership, but also in retention of ownership. In addition to replicating previous findings that transfers increase the probability of becoming a homeowner and that parental wealth is associated with sustained ownership \citep[see e.g.,][]{Lee2018,bond2021role}, I provide novel results on the mechanism through which parental wealth aids retention. Controlling for a wide set of household and parental characteristics, I show that households with wealthier parents are less likely to fall behind on mortgage payments. Finally, using a variation of the event study in \cite{Chetty2007}, I show that while households with less wealthy parents downsize their housing during unemployment spells, those with wealthy parents retain their current homes.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the homeownership rate among young households. In one adaption of my model, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults, following the  methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}. I find that parental transfers account for 7 percentage points (14\%) of the Black-White homeownership gap.

The interaction between altruistic transfers and illiquid housing generates surprising implications for several widely used housing policy levers, many of which disproportionately benefit households with wealthy parents. Relaxing borrowing constraints---for example, by lowering minimum down payments---strengthens the link between parental wealth and children's housing outcomes, since more households with wealthy parents are on the margin to become homeowners. Nonmortgage policies that lower barriers to entry, such as reducing prices or purchase costs, increase homeownership for all households, though disproportionately among those with wealthier parents. However, increasing liquidity by reducing sales costs weakens the link between parental wealth and children's housing outcomes, as the partial insurance provided by parental wealth becomes less important.

I then study how illiquidity affects the strategic behavior in the absence of commitment. I find that 27\% of young households prefer illiquid housing---due to sales costs---to liquid housing. These households have wealthy parents and use the sales costs as a commitment device to maintain low liquid wealth in the next period, effectively becoming ``wealthy hand-to-mouth.'' Lower liquid wealth implies higher marginal propensities to consume (MPC) and, hence, more parental transfers. As an illustration, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption---implying  a high marginal value of wealth---or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home.

That parental wealth generates a preference for illiquidity is not only a new theoretical mechanism---distinct from explanations like temptation or time inconsistency \citep[e.g.,][]{attanasio2024temptation,laibson1997golden}---but one that helps explain the positive link between parental wealth and successfull homeownership documented earlier. It also aligns with \citet{choukhmane2023benefits}, who find that households with wealthy parents are more likely to invest in illiquid retirement accounts and avoid costly early withdrawals, even after controlling for a wide set of characteristics.

I then examine the implications of transfers and housing on marginal propensities to consume (MPCs). First, the presence of altruistic transfers reduces average MPCs by nearly 30\%, suggesting that standard life-cycle models without parent-child interactions may overstate MPCs. Second, the preference for illiquidity induced by transfers offers a novel explanation for the existence of high-MPC households \cite[e.g.,][]{kaplan2022marginal}; in fact, the entire decline in average MPCs is driven by lower MPCs among the wealthy hand-to-mouth. Third, a common explanation for high MPCs is heterogeneity in preferences---such as impatience \citep[e.g.,][]{aguiar2024hand}---but such behavior is observationally similar to that of households with wealthy, altruistic parents.

@@

\subsubsection{Households with Wealthier Parents Retain Ownership}\label{sec:maintaining}
While most research focuses on the transition into ownership, I now focus on the challenge of retaining it.

@@

\begin{table}[tb]
	\center 
	\begin{threeparttable}
	\caption{Model Estimation}\label{tab:esttable}
		\input{../tabfig/est/esttable.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} Standard errors calculated by estimating the model to 100 bootstrapped empirical samples.
		\end{flushleft}\end{footnotesize}}
\end{table}
\subsection{Nontargeted Moments}\label{sec:externalval}
To validate the model, I show that it also matches nontargeted moments.

@@

\begin{figure}[tb]
	\caption{Event Study: Housing Consumption at Unemployment by Parental Wealth}\label{fig:housinggrowthrates_model}
	\includegraphics[width=0.5\textwidth]{../tabfig/model_housinggrowthpoor_both}%
	\includegraphics[width=0.5\textwidth]{../tabfig/model_housinggrowthrich_both}

@@

\subsubsection{Parental Wealth and Ownership Retention}
Next, I repeat the exercise in Section~\ref{sec:maintaining} (see Figure~\ref{fig:owner_hazard}) for first-time homebuyers, where we saw that (a) parental wealth is associated with more retention, (b) more so than own wealth, and (c) that the role of parental wealth inreases over time horizons. In the model, I replicate the regression as closely as possible. For example, there are no year effects or controls for family size. The model captures these patterns well, though it overpredicts the role of parental wealth at shorter horizons.

\begin{figure}[tb]
	\caption{Homeownership Retention Among First-Time Owners}\label{fig:remain_model}
	\includegraphics[width=0.5\textwidth]{../tabfig/model_remain_own.pdf}%
	\includegraphics[width=0.5\textwidth]{../tabfig/model_remain_parent.pdf}

	{\begin{footnotesize} \textit{Notes:} Lines denote means, and shaded areas indicate 95\% confidence intervals. The light orange area denotes the empirical estimates (Section \ref{sec:maintaining}), and the darker grey area denotes the model simulation. The sample consists of households first-time owners aged 25-45. \end{footnotesize}}
\end{figure}

\subsubsection{Taking Stock on Model Fit}
The model matches targeted moments related to homeownership, transfers, and rent-to-own transitions using realistic preference parameters. It also performs well on untargeted aggregate moments. Perhaps more importantly, it replicates key non-targetted household-level patterns in the associations between parental wealth, downsizing behavior, and ownership retention.

@@

The observed Black-White homeownership gap is $\frac{0.54 - 0.27}{0.54} = 50\%$, compared to 52\% in the model with altruism. Removing altruism reduces the gap to 45\%, implying that parental transfers account for 7 percentage points---or 14\%---of the gap. This is a sizeable effect, especially given that the Black–White homeownership gap has remained largely unchanged since the passage of the Fair Housing Act in 1968. While Black and White households receive transfers at similar rates in the model, transfers are less influential for Black households, who are further from the ownership threshold and receive smaller amounts. These results highlight the importance of racial differences in parental income, which shape both wealth accumulation and transfer capacity.

@@

These results have several implications for the MPC literature. First, model-implied MPCs may be overstated when intra-family transfers are omitted, particularly for households with wealthy parents. Second, a common strategy to match high MPCs is to introduce households with low target wealth—e.g., through impatience \citep{aguiar2024hand}—but such behavior is observationally similar to that of households expecting altruistic support. Third, models that generate high MPCs by introducing illiquid assets with large return premia \citep{kaplan2022marginal} may overlook the fact that homeowners with wealthy parents have access to informal insurance, which lowers their effective MPC.
